---
title: "COMP1013 Analytics Programming — Assignment Autumn 2026"
author: "Rohan Shrestha | Student ID: 22172810"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
geometry: margin=2.5cm
fontsize: 11pt
header-includes:
  - \usepackage{booktabs}
  - \usepackage{float}
  - \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
# -------------------------------------------------------
# Global chunk options — hide code output unless needed
# -------------------------------------------------------
knitr::opts_chunk$set(
  echo    = TRUE,    # show code in report
  warning = FALSE,   # suppress warnings
  message = FALSE,   # suppress package messages
  fig.width  = 7,
  fig.height = 4.5,
  fig.align  = "center"
)
```

---

> **Declaration**
>
> By including this statement, we the authors of this work, verify that:
> We hold a copy of this assignment that we can produce if the original is lost or damaged.
> We hereby certify that no part of this assignment/product has been copied from any other
> student's work or from any other source except where due acknowledgement is made in the
> assignment. No part of this assignment/product has been written/produced for us by another
> person except where such collaboration has been authorised by the subject lecturer/tutor
> concerned. We are aware that this work may be reproduced and submitted to plagiarism
> detection software programs for the purpose of detecting possible plagiarism (which may
> retain a copy on its database for future plagiarism checking). We hereby certify that we
> have read and understand what the School of Computing, Engineering and Mathematics defines
> as minor and substantial breaches of misconduct as outlined in the learning guide for this unit.

---

\newpage

# Data Loading and Preparation

```{r load-libraries}
# Load all required libraries
library(tidyverse)    # data wrangling (dplyr) and visualisation (ggplot2)
library(lubridate)    # date handling — year(), as.Date()
library(knitr)        # kable() for tables
library(kableExtra)   # enhanced table formatting (striped, hover styles)
```

```{r load-data}
# ------------------------------------------------------------------
# Load the three CSV datasets
# Ensure all three files are in the same working directory as this .Rmd
# ------------------------------------------------------------------
players  <- read.csv("players.csv",  stringsAsFactors = FALSE)
games    <- read.csv("games.csv",    stringsAsFactors = FALSE)
sessions <- read.csv("sessions.csv", stringsAsFactors = FALSE)

# Convert signup_date from string to R Date type for date arithmetic
players$signup_date <- as.Date(players$signup_date)
```

```{r inspect-data}
# Quick inspection of each dataset
cat("players  :", nrow(players),  "rows,", ncol(players),  "columns\n")
cat("games    :", nrow(games),    "rows,", ncol(games),    "columns\n")
cat("sessions :", nrow(sessions), "rows,", ncol(sessions), "columns\n")

# Check for NA values in all three datasets
cat("\nNA counts — players:\n");  print(colSums(is.na(players)))
cat("\nNA counts — games:\n");    print(colSums(is.na(games)))
cat("\nNA counts — sessions:\n"); print(colSums(is.na(sessions)))
```

No NA values were found in any of the three datasets, so no imputation is required.
NA handling is still applied defensively throughout each task using `na.rm = TRUE`
and `filter(!is.na(...))` to guard against any unexpected missing values.

\newpage

# Task 1 — Player Engagement Analysis by Experience Level

## Task Explanation

For this task I had to split players into three groups depending on when they signed up. Players who joined before 2023 are Veterans, those who joined during 2023 are Intermediate, and anyone who joined after 2023 is New. Once grouped, I calculated how many players are in each group, what their average session length is in minutes, and what their average score looks like. The idea is to see whether being on the platform longer actually makes a difference to how people play.

## Assumptions

I used signup year as the only means of classifying experience - so someone who signed up in 2021 but barely ever played would still be a Veteran. Any players where the signup year could not be determined were also left out. Every session a player has experienced counts towards the group averages - regardless of when that session occurred.

## Data Wrangling

```{r task1-wrangle}
# Step 1: Classify each player into an experience group based on signup year
players_exp <- players %>%
  mutate(
    signup_year      = year(signup_date),          # extract integer year
    experience_level = case_when(
      signup_year <  2023 ~ "Veteran",             # registered before 2023
      signup_year == 2023 ~ "Intermediate",        # registered in 2023
      signup_year >  2023 ~ "New",                 # registered after 2023
      TRUE               ~ NA_character_           # catch-all for unexpected NAs
    )
  ) %>%
  filter(!is.na(experience_level))                 # remove any unclassified rows

# Step 2: Join players with sessions on player_id (left join keeps all players)
player_sessions <- players_exp %>%
  left_join(sessions, by = "player_id")

# Step 3: Summarise key metrics per experience level
task1_summary <- player_sessions %>%
  group_by(experience_level) %>%
  summarise(
    num_players       = n_distinct(player_id),                        # unique player count
    avg_play_time_min = round(mean(play_time_minutes, na.rm = TRUE), 2), # avg session length
    avg_score         = round(mean(score, na.rm = TRUE), 2)           # avg score per session
  ) %>%
  # Order factor levels logically: most to least experienced
  mutate(experience_level = factor(experience_level,
         levels = c("Veteran", "Intermediate", "New"))) %>%
  arrange(experience_level)
```

## Results Table

```{r task1-table}
# Display results as a formatted kable table
task1_summary %>%
  kable(
    col.names = c("Experience Level", "Number of Players",
                  "Avg Play Time (min)", "Avg Score"),
    caption   = "Player Engagement Metrics by Experience Level",
    booktabs  = TRUE
  ) %>%
  kable_styling(latex_options = c("striped", "hold_position"),
                full_width = FALSE)
```

## Visualisation

```{r task1-plot, fig.cap="Average Play Time by Player Experience Level"}
ggplot(task1_summary,
       aes(x = experience_level, y = avg_play_time_min,
           fill = experience_level)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = avg_play_time_min),
            vjust = -0.4, size = 4.5, fontface = "bold") +
  scale_fill_manual(values = c("Veteran"      = "#2C7BB6",
                               "Intermediate" = "#ABD9E9",
                               "New"          = "#FDAE61")) +
  labs(
    title    = "Average Play Time by Player Experience Level",
    subtitle = "Grouped by year of account signup",
    x        = "Experience Level",
    y        = "Average Play Time (minutes)"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

## Interpretation

The results were a bit surprising as all three groups featured nearly exactly the same amount of playing time per session (around 91-92 minutes) and very similar scoring outcomes (496-501). The veterans of the game do indeed edge out new players in average score - which makes sense given experience - but again the difference is so slight as to be effectively nothing. This does tell me, however, that new players are neither struggling to keep up nor dropping off participation - they are just as engaged as those with longer experience levels. This is actually a good finding for Play Pulse Studios suggesting the platform maintains people’s attention regardless of length of time they've been on it.

\newpage

# Task 2 — Game Genre Analysis

## Task Explanation

This task looked at the popularity of each game genre within the platform. There are five genres - Action, Puzzle, RPG, Sports and Strategy. For each of these I worked out the average play time, average score, total number of sessions and how many unique players played that genre. This provides a picture of which game genres people are drawn to and which don’t receive such love.

## Assumption

To get the genre for each session I had to join the sessions table with the games table using game_id. Any sessions without a matching game (and thus no genre) were dropped. While unique player count is the primary measure of popularity I also looked at play time and session count for a more complete view.

## Data Wrangling

```{r task2-wrangle}
# Step 1: Join sessions with games to get genre for every session
sessions_genres <- sessions %>%
  left_join(games, by = "game_id") %>%
  filter(!is.na(genre))           # drop any sessions without a matched genre

# Step 2: Summarise engagement metrics per genre
task2_summary <- sessions_genres %>%
  group_by(genre) %>%
  summarise(
    avg_play_time      = round(mean(play_time_minutes, na.rm = TRUE), 2),
    avg_score          = round(mean(score, na.rm = TRUE), 2),
    num_sessions       = n(),                         # total session count
    num_unique_players = n_distinct(player_id)        # distinct players per genre
  ) %>%
  arrange(desc(num_unique_players))                   # sort by popularity
```

## Results Table

```{r task2-table}
task2_summary %>%
  kable(
    col.names = c("Genre", "Avg Play Time (min)", "Avg Score",
                  "Sessions", "Unique Players"),
    caption   = "Player Engagement Metrics by Game Genre",
    booktabs  = TRUE
  ) %>%
  kable_styling(latex_options = c("striped", "hold_position"),
                full_width = FALSE)
```

## Visualisation

```{r task2-plot, fig.cap="Number of Unique Players by Game Genre"}
ggplot(task2_summary,
       aes(x = reorder(genre, num_unique_players),
           y = num_unique_players, fill = genre)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = num_unique_players),
            hjust = -0.2, size = 4.2) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title    = "Number of Unique Players by Game Genre",
    subtitle = "Based on all recorded session data",
    x        = "Genre",
    y        = "Number of Unique Players"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

## Interpretation

Puzzle came out on top - it had 3,606 unique players and 6,371 sessions - nearly double what RPG got. Sports and Strategy sat in the middle whereas RPG had both the fewest unique players and sessions. One thing I found interesting was that Strategy ended up with the highest average play time despite not being the most popular genre - people who play seem committed to each session. Sports players scored highest again which makes sense with how competitive those games tend to be. If I were advising the studio I’d look to continue investing in Puzzle content as it has the widest appeal but also dig into what makes Strategy so sticky in order to bring that same quality to other genres struggling for success.

\newpage

# Task 3 — Top Players and Their Behaviour

## Task Explanation

Here I needed to find the 10 players who had amassed the most total time across all their sessions. After getting that list I looked at how often these players come back, how long the average session is and what kind of scores they're putting up. Basically I wanted to understand what the heaviest users of the platform look like.

## Assumptions

I positioned players according to their total cumulative play time - the sum of all their sessions. Any player with NA values for play time I excluded those sessions from the total play time ranking. I took all sessions for the behavioural metrics for each of the top 10 players, with NAs dealt with via na.rm.

## Data Wrangling

```{r task3-wrangle}
# Step 1: Rank all players by total play time and keep the top 10
top10_player_ids <- sessions %>%
  filter(!is.na(play_time_minutes)) %>%       # exclude any NA play times
  group_by(player_id) %>%
  summarise(total_play_time = sum(play_time_minutes)) %>%
  slice_max(total_play_time, n = 10) %>%      # top 10 by cumulative play time
  pull(player_id)                             # extract just the IDs

# Step 2: Filter sessions to top 10 players and compute behavioural metrics
task3_summary <- sessions %>%
  filter(player_id %in% top10_player_ids) %>%
  group_by(player_id) %>%
  summarise(
    total_sessions    = n(),
    avg_play_time_min = round(mean(play_time_minutes, na.rm = TRUE), 2),
    avg_score         = round(mean(score, na.rm = TRUE), 2)
  ) %>%
  arrange(desc(avg_play_time_min))
```

## Results Table

```{r task3-table}
task3_summary %>%
  kable(
    col.names = c("Player ID", "Total Sessions",
                  "Avg Play Time (min)", "Avg Score"),
    caption   = "Behaviour Summary of Top 10 Players by Total Play Time",
    booktabs  = TRUE
  ) %>%
  kable_styling(latex_options = c("striped", "hold_position"),
                full_width = FALSE)
```

## Visualisation

```{r task3-plot, fig.cap="Score Distribution for the Top 10 Players (Boxplot)"}
sessions %>%
  filter(player_id %in% top10_player_ids) %>%
  mutate(player_id = factor(player_id)) %>%    # treat ID as a categorical variable
  ggplot(aes(x = player_id, y = score, fill = player_id)) +
    geom_boxplot(show.legend = FALSE,
                 outlier.colour = "red", outlier.shape = 16, outlier.size = 2) +
    scale_fill_brewer(palette = "Paired") +
    labs(
      title    = "Score Distribution for Top 10 Players",
      subtitle = "Players ranked by cumulative play time; red dots = outliers",
      x        = "Player ID",
      y        = "Score"
    ) +
    theme_minimal(base_size = 13) +
    theme(plot.title  = element_text(face = "bold"),
          axis.text.x = element_text(angle = 45, hjust = 1))
```

## Interpretation

The top 10 players show varied behaviour - yet all are high-engagement users. The very top\
player (ID 3332) sits at 133 mins per session - well above platform average of \~92\
minutes. Session counts range from 9 to 13 showing a frequent return to the platform.

The boxplots reveal some notable variability within scores - players such as 2447 who\
score an average of 617 seem consistently high-performing compared to users such as 3812\
whose average score is 429 yet who put in lengthy sessions. Such outcomes suggest\
that high levels of play time do not always correlate with performance - some play\
learning or exploring modes or play casually over longer periods.

\newpage

# Task 4 — Age Group Behaviour Comparison

## Task Explanation

This task divides all players into four age brackets and compares their gaming behaviour. The goal is to determine whether age impacts either how actively or skillfully players interact with this platform. The age groups are:

- **16–25**: Youngest players
- **26–35**: Young adults
- **36–45**: Middle-aged players
- **46+**: Older players

For each group we calculate average play time, average score, and number of sessions.

## Assumptions

- Each player's age is taken as their current age at the time of analysis.

- Players with missing age values are left out.
- All sessions contribute to the statistics of that age group player's statistics.

## Data Wrangling

```{r task4-wrangle}
# Step 1: Assign each player to an age bracket
players_age <- players %>%
  filter(!is.na(age)) %>%           # exclude players with missing age
  mutate(
    age_group = case_when(
      age >= 16 & age <= 25 ~ "16-25",
      age >= 26 & age <= 35 ~ "26-35",
      age >= 36 & age <= 45 ~ "36-45",
      age >= 46              ~ "46+",
      TRUE                   ~ NA_character_   # catch unexpected values
    )
  ) %>%
  filter(!is.na(age_group))         # remove any unclassified players

# Step 2: Join age groups with sessions to attach age info to each session
sessions_age <- players_age %>%
  select(player_id, age_group) %>%
  inner_join(sessions, by = "player_id")

# Step 3: Summarise gaming behaviour per age group
task4_summary <- sessions_age %>%
  group_by(age_group) %>%
  summarise(
    avg_play_time_min = round(mean(play_time_minutes, na.rm = TRUE), 2),
    avg_score         = round(mean(score, na.rm = TRUE), 2),
    num_sessions      = n()
  ) %>%
  # Set factor levels to display age groups in ascending order
  mutate(age_group = factor(age_group,
         levels = c("16-25", "26-35", "36-45", "46+"))) %>%
  arrange(age_group)
```

## Results Table

```{r task4-table}
task4_summary %>%
  kable(
    col.names = c("Age Group", "Avg Play Time (min)",
                  "Avg Score", "Number of Sessions"),
    caption   = "Gaming Behaviour Metrics by Age Group",
    booktabs  = TRUE
  ) %>%
  kable_styling(latex_options = c("striped", "hold_position"),
                full_width = FALSE)
```

## Visualisation

```{r task4-plot, fig.cap="Average Play Time by Age Group"}
ggplot(task4_summary,
       aes(x = age_group, y = avg_play_time_min, fill = age_group)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = avg_play_time_min),
            vjust = -0.4, size = 4.5, fontface = "bold") +
  scale_fill_manual(values = c("16-25" = "#66C2A5",
                               "26-35" = "#FC8D62",
                               "36-45" = "#8DA0CB",
                               "46+"   = "#E78AC3")) +
  labs(
    title    = "Average Play Time by Age Group",
    subtitle = "Comparison of session duration across four age brackets",
    x        = "Age Group",
    y        = "Average Play Time (minutes)"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

## Interpretation

Contrary to the common assumption that younger players are more active the data tells a more subtle story. The 46+ group exhibits both the highest average play time (93.11 min) and the most sessions (6,293) - making this the most active playing cohort. The 16-25 group has the fewest sessions (4,413).

Scores are similar across age groups (\~496-503) indicating no strong differences in skill levels. The 26-35 group scores a marginal highest (503.53).

While gaming activities are commonly thought of as part of a young person’s activities this research shows otherwise. For Play Pulse Studios this indicates that marketing and content development should not neglect older demographic groups - they feature a highly engaged and loyal user base.

\newpage

# Summary

Analysis of gaming data by Play Pulse Studios reveals the following insights:

1. Experience level has minimal influence upon either play time or score - all cohorts of players engage equally within the platform;

2. Puzzle is the dominant genre in terms of popularity (3,606 unique players), while Strategy genres score highest in terms of average play time;

3. Top players are highly diverse - long play times do not necessarily indicate high scoring abilities within games;

4. The most active players are of advancing ages (46+) which disproves the prevailing assumption of such engagement by younger players.
